Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-10-23 mprop/mprop.tex
I noticed I haven't commited in a while...

\documentclass{mprop}
\usepackage{graphicx}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}
\newcommand{\DKL}{D_{\mathrm{KL}}}
\newcommand{\pfull}{p(\mathcal{D}, \Theta, \mathbf{X_u}, \mathbf{u}, \mathbf{r})}
\newcommand{\posterior}{p(\Theta, \mathbf{u}, \mathbf{r} | \mathcal{D}, \mathbf{X_u})}
\newcommand{\approximation}{q(\Theta, \mathbf{u}, \mathbf{r}; \Lambda)}

@@

% TODO: import info from the MSci paper
As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\begin{equation} \label{full}
  \pfull = p(\mathbf{X_u}) \times p(\Theta | \mathbf{X_u}) \times p(\mathbf{u}
  | \Theta, \mathbf{X_u}) \times p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}) \times p(\mathcal{D} | r).
\end{equation}
Here $p(\mathbf{X_u})$ and $p(\Theta | \mathbf{X_u})$ are freely chosen priors,
\[
  \begin{split}
    p(\mathbf{u} | \Theta, \mathbf{X_u}) &= \mathcal{N}(\mathbf{0},
    \Kuu) \\
    &= \frac{1}{(2\pi)^{n/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2}
      \mathbf{u}^T\Kuu^{-1}\mathbf{u} \right) \\
    &= \exp\left(-\frac{1}{2}\mathbf{u}^T\Kuu^{-1}\mathbf{u} -
      \frac{1}{2}\log|\Kuu| - \frac{n}{2}\log 2\pi\right)
  \end{split}
\]
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, the GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\Kru^T\Kuu^{-1}\mathbf{u}, \Krr - \Kru^T\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in Equation \ref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k_\Theta(\mathbf{x}_{\mathbf{r},i}, \mathbf{x}_{\mathbf{u},j})$,
where $\mathbf{x}_{\mathbf{r},i}$ and $\mathbf{x}_{\mathbf{u},j}$ denote feature
vectors for the $i$th state in $\mathcal{S}$ and the $j$th state in
$\mathbf{X_u}$, respectively.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\Theta$, inducing rewards $\mathbf{u}$, and the reward function $r$. While the
previous paper that considered this IRL model computed maximum likelihood estimates
for $\Theta$ and $\mathbf{u}$, and made an assumption that $\mathbf{r}$ in
Equation \ref{eq:r} has $0$ variance \cite{DBLP:conf/nips/LevinePK11}, we aim to
avoid this assumption and use variational inference to approximate the full
posterior distribution $\posterior$. \emph{Variational inference} (VI) is an
approximation technique for probability densities \cite{blei2017variational}.
Let $\approximation$ be our approximating family of probability distributions
for $\posterior$ with its own set of hyperparameters $\Lambda$. Then it is up to
VI algorithms to optimise $\Lambda$ in order to minimise the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are and in this case can be defined as follows
\cite{blei2017variational}:
\[ \DKL(\approximation || \posterior) = \mathbb{E}_q[\log\approximation] -
  \mathbb{E}_q[\log \posterior]. \]
Since KL divergence is typically hard to compute, instead of minimising it, VI
typically tries to maximise the \emph{evidence lower bound} (ELBO) defined as
\[
  \begin{split}
    \mathcal{L}(\Lambda) &= \mathbb{E}_q[\log \pfull] -
    \mathbb{E}_q[\log\approximation] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\Theta\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\]
% TODO: cite something

@@

\section{Literature Survey} \label{literature}

present an overview of relevant previous work including articles, books, and
existing software products. Critically evaluate the strengths and weaknesses of
the previous work.

\section{Proposed Approach}

@@

We begin by rewriting the posterior by using the chain rule and Bayes' theorem,
trying to extract parts of the distribution we know how to compute, namely those
in Equation \ref{full}:
\[
  \begin{split}
    \posterior &= p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} |
    \Theta, \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}, \mathcal{D}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} | r) \times p(\mathbf{r} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}) \times p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\mathcal{D} | \Theta, \mathbf{X_u}) \times p(\Theta |
    \mathbf{X_u}) \times p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u})
    \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \times p(\mathcal{D} | r) \times
    p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u})
  \end{split}
\]
Note that now there are only two unknown probability distributions:
$p(\mathcal{D} | \Theta, \mathbf{X_u})$ and $p(\mathcal{D} | \Theta,
\mathbf{X_u}, \mathbf{u})$. They can be computed as follows:
\begin{align*}
  p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u}) = \int &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \, d\mathbf{r}, \\
  p(\mathcal{D} | \Theta, \mathbf{X_u}) = \iint &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \, d\mathbf{u} \, d\mathbf{r}.
\end{align*}

@@

Spectral kernels \cite{pmlr-v28-wilson13}

@@

Normalizing flows \cite{DBLP:conf/icml/RezendeM15}

@@

Structured stochastic VI (haven't read) \cite{DBLP:conf/aistats/HoffmanB15}

@@

Sparse GPs \cite{DBLP:journals/jmlr/CandelaR05}.

Tighter ELBOs are not necessarily better \cite{DBLP:conf/icml/RainforthKLMIWT18}.

\cite{DBLP:books/lib/Bishop07}

>>>>

\documentclass{mprop}
\usepackage{graphicx}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{url}
\usepackage{bm}
\usepackage{bbm}
\usepackage{booktabs}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\diag}{diag}

@@

% TODO: table of notation?
\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}
\newcommand{\DKL}{D_{\mathrm{KL}}}
\newcommand{\pfull}{p(\mathcal{D}, \Theta, \mathbf{X_u}, \mathbf{u}, \mathbf{r})}
\newcommand{\pfullS}{p(\mathcal{D}, \Theta_s, \mathbf{X_u}, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\posterior}{p(\Theta, \mathbf{u}, \mathbf{r} | \mathcal{D}, \mathbf{X_u})}
\newcommand{\approximation}{q_{\bm\nu}(\Theta, \mathbf{u}, \mathbf{r})}
\newcommand{\approximationS}{q_{\bm\nu}(\Theta_s, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\Eq}{\mathbb{E}_{(\Theta, \mathbf{u}, \mathbf{r}) \sim \approximation}}

@@

% TODO: import info from the MSci paper
As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\begin{equation} \label{full}
  \pfull = p(\mathbf{X_u}) \times p(\Theta | \mathbf{X_u}) \times p(\mathbf{u}
  | \Theta, \mathbf{X_u}) \times p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}) \times p(\mathcal{D} | r).
\end{equation}
Here $p(\mathbf{X_u})$ and $p(\Theta | \mathbf{X_u})$ are freely chosen priors,
\begin{align}
  p(\mathbf{u} | \Theta, \mathbf{X_u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \nonumber \\
                                       &= \frac{1}{(2\pi)^{n/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^T\Kuu^{-1}\mathbf{u} \right) \nonumber \\
                                       &= \exp\left(-\frac{1}{2}\mathbf{u}^T\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right) \label{eq:normal}
  \end{align} % TODO: introduce m here, deleting later introduction
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, the GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^T\Kuu^{-1}\mathbf{u}, \Krr - \Kru^T\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in \eqref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k_\Theta(\mathbf{x}_{\mathbf{r},i}, \mathbf{x}_{\mathbf{u},j})$,
where $\mathbf{x}_{\mathbf{r},i}$ and $\mathbf{x}_{\mathbf{u},j}$ denote feature
vectors for the $i$th state in $\mathcal{S}$ and the $j$th state in
$\mathbf{X_u}$, respectively \cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\Theta$, inducing rewards $\mathbf{u}$, and the reward function $r$. While the
previous paper that considered this IRL model computed maximum likelihood estimates
for $\Theta$ and $\mathbf{u}$, and made an assumption that $\mathbf{r}$ in
\eqref{eq:r} has zero variance \cite{DBLP:conf/nips/LevinePK11}, we aim to
avoid this assumption and use variational inference to approximate the full
posterior distribution $\posterior$. \emph{Variational inference} (VI) is an
approximation technique for probability densities \cite{blei2017variational}.
Let $\approximation$ be our approximating family of probability distributions
for $\posterior$ with its own hyperparameter vector $\bm\nu$. Then it is up
to VI algorithms to optimise $\bm\nu$ in order to minimise the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are and in this case can be defined as follows
\cite{blei2017variational}:
\[ \DKL(\approximation || \posterior) = \Eq \left[ \log
  \frac{\approximation}{\posterior} \right]. \]
Since KL divergence is typically hard to compute, instead of minimising it, VI
typically tries to maximise the \emph{evidence lower bound} (ELBO) defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\[
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\Theta\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\]

@@

\section{Literature Survey} \label{literature}
%present an overview of relevant previous work including articles, books, and
%existing software products. Critically evaluate the strengths and weaknesses of
%the previous work.

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as neural networks)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, while
providing a time complexity of $\mathcal{O}(m^3)$, where $m$ is the number of
inducing points. Below we provide a short overview of various assumptions that
have been used in approximating \emph{sparse} GPs (i.e., GPs that use inducing
points), following on a paper by Qui\~{n}onero-Candela and Rasmussen
\cite{DBLP:journals/jmlr/CandelaR05}.

\begin{description}
  \item [Subset of data ($\mathcal{O}(m^3)$)] is a baseline method of simply
    using a subset of data points.
  \item [Subset of regressors ($\mathcal{O}(nm^2)$)
    \cite{silverman1985some,DBLP:conf/nips/SmolaB00,DBLP:conf/nips/WahbaLGXKK98}]
    is a degenerate approximation that uses a weight for each inducing point. A
    GP is called \emph{degenerate} if the covariance function has a finite
    number of non-zero eigenvalues, restricting the prior distribution to only a
    finite number of linearly independent functions
    \cite{DBLP:journals/jmlr/CandelaR05}.
  \item [Deterministic training conditional ($\mathcal{O}(nm^2)$)
    \cite{DBLP:conf/aistats/SeegerWL03}] approximation imposes a zero-variance
    normal distribution for $\mathbf{r} | \mathbf{u}$, resulting in the same mean
    but different variance predictions compared to the subset of regressors.
  \item [Fully independent training conditional ($\mathcal{O}(nm^2)$)
    \cite{DBLP:conf/nips/SnelsonG05}] has the assumption that the GP values are
    independent of each other when conditioned on the inducing values:
    \[ q(\mathbf{r} | \mathbf{u}) = \prod_{i=1}^n p(r_i | \mathbf{u}). \]
  \item [Partially independent training conditional ($\mathcal{O}(nm^2)$)
    \cite{DBLP:journals/neco/Tresp00,DBLP:conf/nips/SchwaighoferT02}]
    approximates the same distribution as the fully independent training
    conditional, but considers a block diagonal rather than a diagonal
    covariance matrix.
  \item [Transduction] tailors the predictive distribution to
    specific test inputs \cite{DBLP:journals/jmlr/CandelaR05}. As we are not too
    concerned about a specific set of test inputs in the IRL setting,
    transduction is of limited interest to us.
  \item [Augmentation \cite{rasmussen2002reduced}] aims to improve predictive
    accuracy by adding each test input to the inducing points.
  \item [Nystr\"{o}m approximation ($\mathcal{O}(nm^2)$)
    \cite{DBLP:conf/nips/WilliamsS00}] approximates the prior covariance of
    $\mathbf{r}$, but can lead to negative predictive variances.
  \item [Relevance vector machine ($\mathcal{O}(m^3)$)
    \cite{DBLP:journals/jmlr/Tipping01}] is a degenerative approximation
    supporting a limited range of covariance functions.
\end{description}

\begin{table}
  \centering % TODO: recheck and explain
  \begin{tabular}{llllll} % Lower bound? Model? % TODO: comment about 'modelled' no longer needed
    \toprule
    Authors, year & Inducing points & Hyperparameters & Complexity \\
    \midrule
    Titsias, 2009 \cite{DBLP:journals/jmlr/Titsias09} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hensman et al., 2013 \cite{DBLP:conf/uai/HensmanFL13} & fixed & variational & $\mathcal{O}(m^3)$ \\
    Gal et al., 2014 \cite{DBLP:conf/nips/GalWR14} & variational & variational & $\mathcal{O}(nm^2)$ \\
    Hoang et al., 2015 \cite{DBLP:conf/icml/HoangHL15} & fixed & fixed & $\mathcal{O}(m^3)$ \\
    Cheng and Boots, 2017 \cite{DBLP:conf/nips/ChengB17} & variational & variational & $\mathcal{O}(nm_\alpha + nm_\beta^2)$ \\
    Hensman et al., 2017 \cite{DBLP:journals/jmlr/HensmanDS17} & fixed & variational & $\mathcal{O}(nm)$ \\
    Peng et al., 2017 \cite{DBLP:conf/icml/PengZZQ17} & variational & variational & $\mathcal{O}(m^3)$ \\
    \bottomrule
  \end{tabular}
  \caption{Summary of relevant VI approximations to GPs. For both inducing
    points and hyperparameters, 'fixed' means 'chosen before the algorithm
    starts', 'modelled' means 'represented by a probability distribution', and
    'variational' means 'included amongst the variational parameters'.
    Hyperparameters $m_\alpha$ and $m_\beta$ refer to the number of bases used
    to represent the GP's mean and covariance, respectively.}
  \label{table:approximations}
\end{table}

Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be suitable for our GP IRL
model. In order to derive a reliable ELBO, the inducing points should be either
fixed or modelled, while hyperparameters cannot be variational (the reason will
be explained in Section \ref{sec:proposed_approach}).

\begin{itemize}
\item We would also like to avoid having the hyperparameters fixed, as learning
  them efficiently would introduce a separate problem. As the approach by Hoang
  et al. does not seem to be easily extendable to modelled hyperparameters, it
  is unsuitable to our needs.
\item (TODO: this will have comments about other papers as well and will be
  restructured into a paragraph) The variational Fourier features (VFF)
  algorithm by Hensman et al. \cite{DBLP:journals/jmlr/HensmanDS17} is only
  defined for Mat\'ern kernels, which is likely to be too restrictive for our
  situation (the original paper on using GPs for IRL
  \cite{DBLP:conf/nips/LevinePK11} used the automatic relevance detection kernel
  that has weights controlling how important each feature is). While extending
  VFF to support a flexible class of kernels defined by Wilson and Adams
  \cite{pmlr-v28-wilson13} is an interesting and promising avenue of work, it is
  likely to be beyond the scope of this project.
\item The derivation of the ELBO in the work of Peng et al.
  \cite{DBLP:conf/icml/PengZZQ17} relies primarily on the fact that the evidence
  for a GP is a Gaussian, while in our case the evidence can be defined in
  several ways depending on the model, but must always involve $\mathcal{D}$,
  making the main idea of the paper inapplicable to IRL.
\end{itemize}

Since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing more than just a unimodal Gaussian.
The primary way to represent complex posteriors in VI is by using
\emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Unfortunately, this
parametrisation also means that the gradient of the joint probability
distribution w.r.t. variational parameters $\nabla_{\bm\nu}\pfull$ is no
longer zero, making an analytic expression for ELBO impossible using the usual
methods. % TODO: not full (I think) + discuss the possibility of making it
         % variational

\section{Proposed Approach} \label{sec:proposed_approach}

@@

In order to properly investigate the difference between variational inference
and maximum likelihood estimation for the model, we keep other parts of the
model the same. Namely, we set the covariance function to a version of the
automatic relevance detection kernel
\cite{DBLP:conf/nips/LevinePK11,neal2012bayesian}
\[ k_\Theta(\mathbf{x}_i, \mathbf{x}_j) = \zeta\exp\left(
    -\frac{1}{2}(\mathbf{x}_i - \mathbf{x}_j)^T\bm\Lambda(\mathbf{x}_i -
    \mathbf{x}_j) - \mathbbm{1}[i \ne j]\sigma^2\Tr[\bm\Lambda] \right), \]
where $\zeta$ is the overall ``scale'' factor for how similar or distant the
states are,
\[ \bm\Lambda = \diag(\lambda_1, \dots, \lambda_d) = \diag(\bm\lambda) \]
is a diagonal matrix that determines how relevant each feature is (where $d$
denotes the number of features), $\mathbbm{1}$ is defined as
\[ \mathbbm{1}[b] = \begin{cases}
    1 & \mbox{if $b$ is true} \\
    0 & \mbox{otherwise},
  \end{cases} \]
and $\sigma^2$ is set to $10^{-2}/2$ (as the original paper noted that
the value makes little difference to the performance of the algorithm
\cite{DBLP:conf/nips/LevinePK11}). Thus we can set $\Theta = \{ \zeta,
\bm\lambda \}$. Similarly, we keep the expression for the prior of $\Theta$:
\[ p(\Theta | \mathbf{X_u}) = \exp\left( -\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1) \right). \]

Next, we can rewrite the posterior by using the chain rule and Bayes' theorem in
order to get a better sense of what we are trying to approximate:
\[
  \begin{split}
    \posterior &= p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} |
    \Theta, \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}, \mathcal{D}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} | r) \times p(\mathbf{r} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\Theta | \mathbf{X_u}, \mathcal{D}) \times p(\mathcal{D} |
    \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta,
    \mathbf{X_u}) \times p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta,
    \mathbf{X_u}, \mathbf{u}) \\
    &\propto p(\mathcal{D} | \Theta, \mathbf{X_u}) \times p(\Theta |
    \mathbf{X_u}) \times p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u})
    \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \times p(\mathcal{D} | r) \times
    p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u})
  \end{split}
\]
Note that now there are only two unknown probability distributions:
$p(\mathcal{D} | \Theta, \mathbf{X_u})$ and $p(\mathcal{D} | \Theta,
\mathbf{X_u}, \mathbf{u})$, which can be computed as follows:
\begin{align*}
  p(\mathcal{D} | \Theta, \mathbf{X_u}, \mathbf{u}) = \int &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \, d\mathbf{r}, \\
  p(\mathcal{D} | \Theta, \mathbf{X_u}) = \iint &p(\mathcal{D} | r) \times p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) \times p(\mathbf{u} | \Theta, \mathbf{X_u}) \, d\mathbf{u} \, d\mathbf{r}.
\end{align*}
This suggests the following form for the approximation:
\[ \approximation = q(\Theta) \times q(\mathbf{u} | \Theta) \times q(\mathbf{r}
  | \Theta, \mathbf{u}). \]

%where $q(\Theta)$ can be of the same form as $p(\Theta | \mathbf{X_u})$,
%$q(\mathbf{u} | \Theta) = \mathcal{N}(\mathbf{0}, \bm\Sigma_{\mathbf{u}})$, and
%$q(\mathbf{r} | \Theta, \mathbf{u}) = \mathcal{N}(\bm\mu_{\mathbf{r}},
%\bm\Sigma_{\mathbf{r}})$.

\subsection{The Structure of the Approximating Distribution}

At this point we are forced to make assumptions about the approximate posterior
in order to arrive at an implementable solution. Ways to relax the assumptions
may be investigated towards the end of the project, if time permits.

First, as is common in the literature for applying VI to GPs
\cite{DBLP:conf/nips/ChengB17,DBLP:conf/uai/HensmanFL13,DBLP:conf/icml/HoangHL15,DBLP:journals/jmlr/Titsias09},
we simply set
\[ q(\mathbf{r} | \Theta, \mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}). \]
We can make a similarly justified choice for $q(\mathbf{u} | \Theta)$:
\[ q(\mathbf{u} | \Theta) = q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S}) \]
for some variational parameters $\mathbf{m} \in \mathbb{R}^{m}$ and $\mathbf{S}
\in \mathbb{R}^{m \times m}$
\cite{DBLP:conf/nips/ChengB17,DBLP:journals/jmlr/HensmanDS17,DBLP:conf/aaai/HoangHL17}.

In order to have a reasonable way of calculating/approximating the ELBO (see
Section \ref{vi_algs}), we need to have an approximating distribution for
$\Theta$, but unfortunately all the papers in Table \ref{table:approximations}
either fix it or treat it as a variational parameter. Hence we make our first
assumption without justification from previous literature:
\[ q(\Theta) = q(\zeta) \times q(\bm\lambda). \]
As it is reasonable for $\bm\lambda$ to take any value in $\mathbb{R}^d$, we can
approximate it simply as
\[ q(\bm\lambda) = \mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma) \]
for some variational parameters $\bm\mu \in \mathbb{R}^d$ and $\bm\Sigma \in
\mathbb{R}^{d \times d}$. In contrast, as the covariance function must produce
non-negative values, we want to restrict $\zeta$ to positive real numbers. At
the same time, the distribution for $\zeta$ should have a flexible mean (i.e.,
not be tied to zero, like in the exponential distribution) and tails that
converge to zero as the value of the random variable moves away from the mean.
We might want to support some right skew, but the distribution should be close
to symmetric with at least some parameter values. This limits our choice of
distributions quite significantly, and we decide to go with the gamma
distribution as it is fairly flexible and commonly used
\cite{hogg2018introduction}. We then define the probability density function of
the gamma distribution as
\[ q(\zeta) = \Gamma(\zeta; \alpha, \beta) =
  \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta}, \]
where $\alpha > 0$ and $\beta > 0$ are two parameters of the distribution, and
$\Gamma(\cdot)$ is the gamma function. This gives us our list of variational
parameters: $\mathbf{m}, \mathbf{S}, \bm\mu, \bm\Sigma, \alpha, \beta$.

\subsection{Evidence Lower Bound}

In this section we derive and simplify the ELBO for this (now fully specified)
model. We begin with a lemma that will be useful during the derivation.

\begin{lemma}
  Let $\mathbf{x} \sim \mathcal{N}(\bm\mu, \bm\Sigma)$ be a random Gaussian
  vector with $n$ elements, and let $f$ be its probability density function.
  Then
  \[ \mathbb{E}[\log f(\mathbf{x})] = -\frac{1}{2}\log |\bm\Sigma| + c, \]
  where $c \in \mathbb{R}$ is a constant w.r.t. $\bm\mu$ and $\bm\Sigma$.
\end{lemma}

\begin{proof} % TODO: cite the matrix cookbook
  Similarly to \eqref{eq:normal}, we can write
  \[ \mathbb{E}[\log f(\mathbf{x})] = \mathbb{E}\left[ -\frac{1}{2}(\mathbf{x} -
      \bm\mu)^T\bm\Sigma^{-1}(\mathbf{x} - \bm\mu) - \frac{1}{2}\log |\bm\Sigma|
      - \frac{n}{2}\log 2\pi \right]. \]
  Note that the term
  \[ \mathbb{E}\left[ -\frac{n}{2}\log 2\pi \right] = -\frac{n}{2}\log 2\pi \]
  can be absorbed into $c$. Now
  \[ \mathbb{E}[(\mathbf{x} - \bm\mu)^T\bm\Sigma^{-1}(\mathbf{x} - \bm\mu)] =
    \Tr(\bm\Sigma^{-1}\mathbf{M}) + \mathbf{c}^T\bm\Sigma^{-1}\mathbf{c}, \]
  where $\mathbf{M} = \Var[\mathbf{x} - \bm\mu]$ and $\mathbf{c} =
  \mathbb{E}[\mathbf{x} - \bm\mu]$ \cite{petersen2008matrix}. But
  \[ \mathbf{c} = \mathbb{E}[\mathbf{x} - \bm\mu] = \mathbb{E}[\mathbf{x}] -
    \bm\mu = \bm\mu - \bm\mu = \mathbf{0}, \]
  so
  \[ \mathbb{E}[(\mathbf{x} - \bm\mu)^T\bm\Sigma^{-1}(\mathbf{x} - \bm\mu)] =
    \Tr(\bm\Sigma^{-1}\mathbf{M}) = \Tr(\mathbf{I}), \]
  since
  \[ \mathbf{M} = \Var[\mathbf{x} - \bm\mu] = \Var[\mathbf{x}] = \bm\Sigma. \]
  Thus, since we have that
  \[ \mathbb{E}\left[ -\frac{1}{2}\log |\bm\Sigma| \right] = -\frac{1}{2}\log
    |\bm\Sigma|, \]
  we get that
  \[ \mathbb{E}[\log f(\mathbf{x})] = -\frac{1}{2}\log |\bm\Sigma| + c \]
  for some $c \in \mathbb{R}$ that depends on $n$, but not on $\bm\mu$ or
  $\bm\Sigma$.
\end{proof}

\subsection{Variational Inference Algorithms} \label{vi_algs}

% TODO: will need to make this equation more precise, once I know what
% approximation I want to use
The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient \cite{blei2017variational}. Due to terms involving
$\mathcal{D}$ that are computed by solving an MDP and do not resemble a typical
probability distribution, we turn our attention to the \emph{black box
  variational inference} \cite{DBLP:conf/aistats/RanganathGB14} paper that
suggests a convenient trick:
\[ \nabla_{\bm\nu} \mathcal{L} = \Eq[\nabla_{\bm\nu} \log \approximation
  (\log \pfull - \log \approximation)]. \]
With this trick in mind, we only need to evaluate $\log \pfull$ and take the
gradient of $\log \approximation$. Following the same paper,
$\nabla_{\bm\nu} \mathcal{L}$ then has an unbiased estimate
\[ \nabla_{\bm\nu} \mathcal{L} \approx \frac{1}{S} \sum_{s=1}^S
  \nabla_{\bm\nu} \log \approximationS (\log \pfullS - \log
  \approximationS) \]
computed by drawing $S$ Monte Carlo samples $(\Theta_s, \mathbf{u}_s,
\mathbf{r}_s) \sim \approximation$.

@@

@@

Structured stochastic VI \cite{DBLP:conf/aistats/HoffmanB15}

Tighter ELBOs are not necessarily better \cite{DBLP:conf/icml/RainforthKLMIWT18}

For details on Lebesgue's dominated convergence theorem \cite{cinlar_probability}

Still looking for the relevant version of the theorem \cite{knill2017probability}

Approximation as a multivariate Gaussian (haven't read) \cite{DBLP:journals/sac/TanN18}

How black box VI can be even more black box \cite{li2016wild}

Evaluating VI \cite{DBLP:conf/icml/YaoVSG18}

\subsubsection{for GPs}

@@

@@

Stochastic VI for sparse spectral GPs (no inducing points, so not that relevant) \cite{DBLP:conf/aaai/HoangHL17}

Sparse GPs 2 \cite{DBLP:conf/nips/GalWR14}

The 3 papers I need to focus on:

SVI for sparse GPs \cite{DBLP:conf/uai/HensmanFL13}

distributed 1 \cite{DBLP:conf/icml/HoangHL15}

distributed 2 (haven't read) \cite{DBLP:conf/icml/PengZZQ17}

The paper at the core of the 3 approaches (approximates posterior) \cite{DBLP:journals/jmlr/Titsias09}

Generalized version? \cite{DBLP:conf/icml/ShethWK15}
